Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-05 doc/papers/cgo2014/paper.tex
Additional rework of the paper

The limitation of expression templates presented in
\S~\ref{sec:expression-templates} can be solved using forward substitution at
the Python compiler level. This optimization leverages the
\textit{def-use} chains as well as interprocedural read-write
effects and alias analysis presented in the next sections.
Thanks to these analyses, the forward substitution implementation is quite
simple: \todo{it is not THAT simple indeed, there is "lazyness" analysis
involved}....

This optimization is able to forward \texttt{t0} and \texttt{t1} in
Listing~\ref{lst:rosen}, avoiding a temporary allocation and providing more
arithmetic intensity at the same time.

@@

Alias analysis is impossible for Python programs in general, because of late
binding and possible monkey patching---a way to modify the code of dynamic
languages at run-time. In the context of a static compiler for
numerical programs, it can be assumed that binding is done statically and no
monkey patching is done. Indeed the Pythran compiler operates on a subset of
Python that forbid this kind of dynamic patching.
In this context, it becomes possible to design an intra-procedural alias
analysis thanks to the absence of pointer arithmetic:
the only language construction that can create aliasing is the assignment, thus
a simple control-flow based alias analysis yields accurate results.

\subsection{Interprocedural Effects Analysis}
\label{sec:effects}

@@

To interprocedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions, and with as many edges as formal parameters between two nodes. Local
variables used as arguments do not trigger an edge.
Edges are labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The analysis is iteratively pruning all edges from $G$ by considering any
node for which the intraprocedural analysis shown that an argument $i$ is
written. Since the parameter is modified, any outgoing edge $(i, j)$ for this
node will not add any information to the analysis and then can be safely
removed. Incoming edges $(j, i)$ are used to propagate the effects along the
call graph: $j$ is added to the source's list of argument written and the source
node is added to the list of nodes waiting to be processed. Iteratively
applying this process removes all edges that can propagate a write, when no edge
can be removed, the remaining ones propagate a read and can be ultimately
removed.

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its arguments or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially feasible in Python thanks to
the ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

Once interprocedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may be exposed, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna} only.

@@

    \centering{
\begin{tabular}{|l||c|c|c|c|c|}
\hline
& Python & numba & parakeet & pypy & pythran \\
\hline
\hline
         allpairs\_distances &   58293 &             0 &          3170 &   59232 & \textbf{1583} \\
  allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
               arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
                        conv & 3260079 &             0 & \textbf{3284} & 3268298 &          3420 \\
                create\_grid &    5231 & \textbf{5044} &             0 &    5240 &          5587 \\
                    cronbach &    1905 &          2016 &             0 &    1909 & \textbf{1890} \\
                   diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
                      evolve &    8236 &             0 &             0 &    8319 & \textbf{6242} \\
                        fdtd & 1940884 &       1498849 &             0 & 2007839 & \textbf{2294} \\
                         fft &   27120 &             0 &             0 &   27009 & \textbf{2151} \\
                    grouping &    3739 &             0 &             0 &    3740 & \textbf{1747} \\
                     growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
                      harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
                     hyantes &  288852 &             0 & \textbf{2834} &  289074 &          3522 \\
                       julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
                      l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
               local\_maxima &  132237 &             0 &             0 &  132047 & \textbf{8561} \\
                      lstsqr &    9254 &          7528 &             0 &    8992 & \textbf{1414} \\
                      mandel &  951612 & \textbf{8175} &             0 &  951687 &         10341 \\
               multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
                    pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
              periodic\_dist &    2559 &          2675 &             0 &    2562 & \textbf{2292} \\
                   repeating &    2808 &          2836 &             0 &    2795 &  \textbf{810} \\
             reverse\_cumsum &    4583 &          4828 &             0 &    4583 & \textbf{2379} \\
                       rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
                   slowparts &    8302 &             0 & \textbf{1610} &    8241 &          1927 \\
                   smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
             specialconvolve &   12537 &         11978 &          6039 &   12570 & \textbf{5572} \\
                vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
                        wave &   72394 &         67590 &             0 &   72300 & \textbf{5186} \\
                       wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
\hline
\end{tabular}
}
\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
\label{tbl:overview}
\end{table*}

Table~\ref{tbl:overview} presents the average execution time for each kernel
and each compiler, with the lowest values highlighted in
bold font. Zeros are used when the given compiler failed to
compile the kernel. Concerning the Numba compiler, this
benchmarks confirms that it works best on explicit loops. It
gives no performance gain when run on high-level kernels,
either falling back to the generation of calls to Python C
API, or stopping compilation. PyPy is the most robust of
the compilers, but to a few exceptions, it does not generate
code more efficient than the original version. This reflects
the fact that Numpy support is still a work-in-progress.
Parakeet performs extremely well on high-level benchmarks.
Although it does not supports some Numpy constructs, it
successfully parallelizes many of them, and also generates
efficient code for explicit loops. Pythran still performs better
than Parakeet on some kernels, whenever it successfully
vectorizes operations.

@@

@@

The Pythran project is an independent Open Source initiative that has been
partially funded by the CARP Project and the SILKAN Company.

>>>>

The limitation of expression templates presented in
\S~\ref{sec:expression-templates} can be solved using forward substitution at
the Python compiler level. This optimization leverages the
\textit{def-use} chains as well as inter-procedural read-write
effects and alias analysis presented in the next sections.
It is able to forward \texttt{t0} and \texttt{t1} in
Listing~\ref{lst:rosen}, avoiding a temporary allocation and providing more
arithmetic intensity at the same time.

@@

Alias analysis is impossible for Python programs in general, because of late
binding and possible monkey patching---a way to modify the code of dynamic
languages at run-time. In the context of a static compiler for
numerical programs, it can be assumed that binding is done statically and no
monkey patching is done. Indeed the Pythran compiler operates on a subset of
Python that forbid this kind of dynamic patching.
In this context, it becomes possible to design an intra procedural alias
analysis thanks to the absence of pointer arithmetic:
the only language construction that can create aliasing is the assignment, thus
a simple control-flow based alias analysis yields accurate results.

\subsection{Inter-procedural Effects Analysis}
\label{sec:effects}

@@

To inter-procedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions, and with as many edges as formal parameters between two nodes. Local
variables used as arguments do not trigger an edge.
Edges are labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The analysis is iteratively pruning all edges from $G$ by considering any
node for which the intra procedural analysis shown that an argument $i$ is
written. Since the parameter is modified, any outgoing edge $(i, j)$ for this
node will not add any information to the analysis and then can be safely
removed. Incoming edges $(j, i)$ are used to propagate the effects along the
call graph: $j$ is added to the source's list of argument written and the source
node is added to the list of nodes waiting to be processed. Iteratively
applying this process removes all edges that can propagate a write, when no edge
can be removed, the remaining ones propagate a read and can be ultimately
removed.

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its arguments or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially feasible in Python thanks to
the ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{Fibonacci(4)} at compile-time.

@@

Once inter-procedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may be exposed, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna} only.

@@

    \centering{
\begin{tabular}{|l||c|c|c|c|c|}
\hline
& Python & numba & parakeet & pypy & pythran \\
\hline
\hline
         allpairs\_distances &   58293 &             N/A &          3170 &   59232 & \textbf{1583} \\
  allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
               arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
                        conv & 3260079 &             N/A & \textbf{3284} & 3268298 &          3420 \\
                create\_grid &    5231 & \textbf{5044} &             N/A &    5240 &          5587 \\
                    cronbach &    1905 &          2016 &             N/A &    1909 & \textbf{1890} \\
                   diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
                      evolve &    8236 &             N/A &             N/A &    8319 & \textbf{6242} \\
                        fdtd & 1940884 &       1498849 &             N/A & 2007839 & \textbf{2294} \\
                         fft &   27120 &             N/A &             N/A &   27009 & \textbf{2151} \\
                    grouping &    3739 &             N/A &             N/A &    3740 & \textbf{1747} \\
                     growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
                      harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
                     hyantes &  288852 &             N/A & \textbf{2834} &  289074 &          3522 \\
                       julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
                      l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
               local\_maxima &  132237 &             N/A &             N/A &  132047 & \textbf{8561} \\
                      lstsqr &    9254 &          7528 &             N/A &    8992 & \textbf{1414} \\
                      mandel &  951612 & \textbf{8175} &             N/A &  951687 &         10341 \\
               multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
                    pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
              periodic\_dist &    2559 &          2675 &             N/A &    2562 & \textbf{2292} \\
                   repeating &    2808 &          2836 &             N/A &    2795 &  \textbf{810} \\
             reverse\_cumsum &    4583 &          4828 &             N/A &    4583 & \textbf{2379} \\
                       rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
                   slowparts &    8302 &             N/A & \textbf{1610} &    8241 &          1927 \\
                   smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
             specialconvolve &   12537 &         11978 &          6039 &   12570 & \textbf{5572} \\
                vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
                        wave &   72394 &         67590 &             N/A &   72300 & \textbf{5186} \\
                       wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
\hline
\end{tabular}
}
\caption{Average execution times for all kernels of the numpy-benchmarks on Intel8 with OpenMP (8 threads) and SIMD extension.}
\label{tbl:overview}
\end{table*}

Table~\ref{tbl:overview} presents the average execution time for each kernel
and each compiler, with the lowest values highlighted in bold font. N/A are
used when the given compiler failed to compile the kernel. Standard deviations
are not presented in the table, as the result computation were very stable, to
the notable exception of \texttt{periodic\_dist} for parakeet. It comes out
that parakeet leaks the memory of some of the \texttt{ndarray} it allocates,
which results in the kernel swapping.

Concerning the Numba compiler, this benchmarks confirms that it works best on
explicit loops (benchmarks \texttt{growcut}, \texttt{mandel},
\texttt{pariwise}). It does not show any significant improvements on high-level
kernels (\texttt{l2norm}, \texttt{multiple\_sum}\dots), either falling back to
the generation of calls to Python C API, or stopping compilation.

PyPy is the most robust of the compilers, but to a few exceptions, it does not
generate code more efficient than the original version. This reflects the fact
that Numpy support is still a work-in-progress.

Parakeet performs extremely well on both high-level benchmarks and explicit
loops. Although it does not supports some Numpy constructs (e.g. it does no
support calling the \texttt{numpy.cumsum} function in \texttt{reverse\_cumsum})
, it successfully parallelizes many of them. It does not vectorize them though,
which explains while Pythran still performs better by more than a factor of
$\times2$ on some kernels.

Pythran outperforms the other compilers, sometimes by a factor greater than
$\times2$, thanks to the combination of vectorization and parallelization. Code
generation for explicit loops can still be improved, as shown by the poor
performance on the \texttt{growcut} kernel and \texttt{hyantes} kernels.

@@

This paper presents a study of the optimization of Python/Numpy high level
kernels in the context of high performance computing. It uses a set of
realistic benchmarks adapted from the question and answers on the StackOverflow
website to identify optimization opportunities. It focuses on the efficient
implementation and optimization of array expressions within the ahead-of-time
Pythran compiler, showing that a compilation step at Python level before
generation of lower-level code makes it possible to generate vectorized,
parallel C++ code. A comparison with existing JIT compiler for scientific
Python validates the approach, showing significant performance improvements
over the state of the art.

@@

The Pythran project is an independent Open Source initiative partially funded
by the CARP Project and the SILKAN Company.
